---
title: "Count Models"
subtitle: "Poisson, Negative Binomial, Zero-Inflation, and Hurdle Models"
date: "`r Sys.Date()`"
author: "Christopher Weber"
organization: "University of Arizona"
email: "chrisweber@arizona.edu"
format: 
  revealjs:
    theme: simple   
    slide-number: true
    toc: false
    toc-depth: 2
    code-fold: true
    code-summary: "Show/Hide Code"
    transition: fade
    center: true
    width: 1200
    height: 800
    slide-level: 2
    navigation-mode: linear
    self-contained: false
    css: styles.css
title-block-banner: "#378DBD"  
---

## Introduction  {.smaller}

-   This lecture follows Long (1997, Chapters 7-8).

-   Many variables in the social sciences consist of integer counts.

 - E.g., number of times a candidate runs for office, frequency of conflict, number of terror attacks, number of war casualties, number of positive statements about a candidate, number of homicides in a city, etc.

-   Least squares is inappropriate for the reasons we've already discussed: $\textbf{Nonlinearity, nonadditivity and heteroskedasticity}$.

-  Instead, we need models that are designed to handle count data.

-  Starting with the most straightforward, let's consider the **Poisson Distribution** and by extension, the **Poisson Regression Model**.

- Then, we'll extend the model to the **Negative Binomial Regression Model** to account for overdispersion.

## The Poisson Distribution  {.smaller}

- The **Poisson Distribution** is governed by one parameter, $\mu$, commonly called the "rate parameter."

- The variance of the **Poisson Distribution** is also governed by $\mu$.

$$p(y|\mu)={{exp(-\mu)\mu^y}\over{y!}}$$

## The Poisson Distribution  {.smaller}

```{r}
plot(dpois(c(0:20),0.12), type="l", main="mu=0.12")
```

##  {.smaller}
```{r}
plot(dpois(c(0:20),1), type="l", main="mu=1")
```

##  {.smaller}
```{r}
plot(dpois(c(0:20),10), type="l", main="mu=10")
```

##  {.smaller}
```{r}
plot(dpois(c(0:20),100), type="l", main="mu=100")
```

## The Poission Distribution  {.smaller}

```{r}
#| echo: false
#| warning: false
#| message: false

library(plotly)

mu_values <- seq(0, 40, by = 1)
data <- expand.grid(x = 0:25, mu = mu_values)
data$prob <- dpois(data$x, data$mu)

p <- plot_ly(data, 
             x = ~x, y = ~prob, frame = ~mu, 
             type = 'scatter', mode = 'lines',
             line = list(width = 3, color = '#378DBD')) %>%
  layout(
    title = list(text = "Poisson Distribution", font = list(size = 18)),
    xaxis = list(title = "Count", range = c(0, 25), titlefont = list(size = 16)),
    yaxis = list(title = "Probability", range = c(0, 0.3), titlefont = list(size = 16)),
    showlegend = FALSE
  ) %>%
  animation_opts(
    frame = 200,
    transition = 150,
    redraw = FALSE
  ) %>%
  animation_slider(
    currentvalue = list(prefix = "Rate Parameter: μ = ", font = list(size = 16, color = "black"))
  )
p
```

## The Poission Distribution  {.smaller}

- As $\mu$ increases, the distribution becomes more symmetric and approaches a normal distribution.

- As $\mu$ increases, the variance increases. A model built from the Poisson Distribution is **heteroskedastic**.

- As $\mu$ decreases, the distribution becomes more right-skewed.

- As $\mu$ decreases, the variance decreases.

## Outline {.smaller}

-   The **Poisson Regression Model** (PRM) is built from the Poisson Distribution. The model allows us to capture the distribution of integer counts, which are typically right skewed.

-   The PRM relies on strong assumption: $E(\mu)=var(y)$. If the data violate this assumption, the variance estimates will be biased, as will test statistics and hypothesis tests.

-   **Overdispersion**. Does the Poisson adequately capture the variance in the data?

-   **Zero Inflation**. Does the Poisson adequately capture the number of zeros in the data?
-    Key Extensions: Negative binomial and zero-inflated models.

## The Poisson Regression Model  {.smaller}

- Absent covariates, the Poisson distribution is given by

$$p(y|\mu)={{exp(-\mu)\mu^y}\over{y!}}$$

-   If $y$ takes on values from 0, 1, 2, 3, etc. (Long 1997, p. 218)
-   In this density, the only parameter that governs the shape of the density is $\mu$ or the "rate" parameter
-   A characteristic of the poisson distribution is that $E(y)=var(y)=\mu$, an assumption called **equidispersion**

## The Poisson regression Model  {.smaller}

-   If $\mu$ is the rate parameter, we can then model $\mu_i$ based on a set of covariates. That is,

$$\mu_i=E(y_i|x_i)=exp(\alpha+\beta x_i)$$

-    $\alpha+\beta x_i$ must be positive; the rate parameter must be positive, hence $exp()$

-   Non-linear regression model with heteroskedasticity (Long 1997, p. 223), $\mu_i=exp(\alpha+\beta x_i)$

-  Note that by extension, the variance is also influenced by covariate(s), $x_i$.

## A Motivating Example  {.smaller}

-   Assume $\alpha=-0.25$, and $\beta=0.13$ as Long does (p. 221). We may calculate the expected values of $y$ for a number of $x$ values, then

-   When $x=1$, then the expected number of counts -- the rate parameter -- is 0.89 (exp(-0.25 + 0.13\* 1))

-   When $x=10$ then the expected number of counts is 2.85 

-    $E(y|x)=exp(\alpha+\beta x_i)=var(y|x)$

## Predicted Probabilities from the Model {.smaller}

- Assume that `mu <- exp(-0.25 + 0.13*1)`

- Then, what is the probability of observing 0, 1, or 2 counts when $x=1$?

```{r}
#| echo: true
#| warning: false
#| message: false

mu <- exp(-0.25 + 0.13*1) 

(exp(-mu) * mu^0) / factorial(0)

(exp(-mu) * mu^1) / factorial(1)

(exp(-mu) * mu^2) / factorial(2)

dpois(0:2, lambda = mu)
```


## The Poisson Regression Model  {.smaller}

-   With $k$ predictors, then

$$\mu_i=exp(\alpha+\sum_K \beta_k x_{k,i})$$

$$p(y|x)={{exp(-exp(\alpha+\sum_K \beta_k x_{k,i}))exp(\alpha+\sum_K \beta_k x_{k,i})^{y_i}}\over{y_i!}}$$
- The Likelihood of the PRM with $k$ predictors:

$$\prod_{i=1}^{N}p(y_i|\mu_i)=\prod_{i=1}^{N}{{exp(-exp(\alpha+\sum_K \beta_k x_{k,i}))exp(\alpha+\sum_K \beta_k x_{k,i})^{y_i}}\over{y_i!}}$$

## The PRM Likelihood  {.smaller}

The log of the likelihood is then,

$$\log\left(\prod_{i=1}^{N}p(y_i|\mu_i)\right) = \sum_{i=1}^{N}\log\left(\frac{\exp\left(-\exp\left(\alpha + \sum_{k=1}^{K} \beta_k x_{k,i}\right)\right) \left[\exp\left(\alpha + \sum_{k=1}^{K} \beta_k x_{k,i}\right)\right]^{y_i}}{y_i!}\right)$$
$$E(y|x)=\mu_i = exp(\alpha+\sum_K \beta_k x_{k,i})$$

## Interpretation {.smaller}

-   The partial derivative of $E(y|x)$ with respect to $x_k$

-   Using the chain-rule.

$$u=\alpha+\sum_K \beta_k x_{k,i}$$ 

$$
\frac{\partial y}{\partial x} = \frac{\partial y}{\partial u} \cdot \frac{\partial u}{\partial x}
$$

## Interpretation {.smaller}

$$
\frac{\partial y}{\partial x} = \frac{\partial y}{\partial u} \cdot \frac{\partial u}{\partial x}
$$

$$ {{\partial E(Y|X)}\over{\partial x_k}}={{\partial exp(u)}\over{\partial u}}{{\partial u \beta}\over{\partial x_k}}$$

$$
exp(\alpha+\sum_K \beta_k x_{k,i})\beta_k=E(Y|X)\beta_k
$$

-  The effect of $x_k$ on $y$ is now a function of the rate parameter **and** the expected effect of $x_k$ on that rate parameter

- It's not a constant change in the expected count; instead it's a function of how $x_k$ affects the rate as well as how all others relate to the rate!

## Interpretation: Change in $x$  {.smaller}

-   What effect does a $d_k$ change in $x_k$ have on the expected count. Take the ratio of the the prediction including the change over the prediction absent the change (Long 1997, p. 225):

$$ {{E(y|X, x_k+d_k)}\over{E(y|X, x_k)}}$$

-   The numerator: 
$$ 
E(y|X, x_k+d_k)=exp(\beta_0)exp(\beta_1 x_1)exp(\beta_2 x_2)...exp(\beta_1 x_k)exp(\beta_k d_k) 
$$

-   The denominator: 
$$
E(y|X, x_k)=exp(\beta_0)exp(\beta_1 x_1)exp(\beta_2 x_2)...exp(\beta_1 x_k)
$$

-  We're left with:  

$$ 
exp(\beta_k d_k)
$$

## Interpretation: Probabilities {.smaller}

-  The predicted probability of a count

$$
pr(y=m|x)={{exp(-exp(\alpha+\sum_K \beta_k x_{k,i}))exp(\alpha+\sum_K \beta_k x_{k,i})^{m}}\over{m!}}
$$

- The partial derivative

$$ exp(\alpha+\sum_K \beta_k x_{k,i})\beta_k=E(Y|X)\beta_k$$

## Summary   {.smaller}

- If $y \sim Poisson(\mu)$, then $E(y)=var(y)=\mu$

- $log(\mu)=\alpha+\sum_K \beta_k x_{k,i}$

-    We might encounter either under or overdispersion brought about by unobserved heterogeneity.

-   It is useful to rely on an alternative model that doesn't treat $\mu$ as fixed, but rather it is drawn from a distribution, i.e., $$\mu_i=exp(\alpha+\sum_K \beta_k x_{k,i})\beta_k+\epsilon_i)$$


## Poisson Regression: Application  {.smaller}

- Scholarly publications of biochemistry graduate students.

- From Long (1997), data consist of 915 biochemistry graduate students and the following observations.

- The data are in `pscl::bioChemists`

- `art`: Number of articles published
- `phd`: Prestige of PhD granting institution 
- `fem`: Whether the student identifies as "Male" or "Female"
- `mar`: Marital status of the student
- `kid5`: Number of children under age 5.
- `ment`: Number of articles published by the mentor during past 3 years.

- We can fit the data using `glm(formula, data, family=poisson(link="log"))`

- Use `pois` family for post-estimation, along with `predict`

- Follow the same workflow: Estimation $\rightarrow$ Prediction $\rightarrow$ (Graphical) Presentation.

## Poisson Regression: Application  {.smaller}

$$
y_{articles} = \beta_0 + \beta_1 x_{kid5} + 
                \beta_2 x_{fem} + \beta_3 x_{mar} + 
                \beta_4 x_{phd} + \beta_5 x_{ment} + 
                \epsilon
$$

```{r}
#| echo: true
#| warning: false
#| message: false

library(pscl)
data(bioChemists)
bioChemists$fem<-as.numeric(bioChemists$fem)-1
bioChemists$mar<-as.numeric(bioChemists$mar)-1
summary(glm(art~kid5 + fem + mar + phd + ment , 
            data=bioChemists,
            family=poisson(link="log")))
```


## Poisson Regression: Application  {.smaller}

$$
y_{articles} = \beta_0 + \beta_1 x_{kid5} + 
                \beta_2 x_{fem} + \beta_3 x_{mar} + 
                \beta_4 x_{phd} + \beta_5 x_{ment} + 
                \epsilon
$$

```{r}
#| echo: false
#| warning: false
#| message: false

library(pscl)
data(bioChemists)
bioChemists$fem<-as.numeric(bioChemists$fem)-1
bioChemists$mar<-as.numeric(bioChemists$mar)-1
my_model = glm(art~kid5 + fem + mar + phd + ment , 
            data=bioChemists,
            family=poisson(link="log"))
```

## Generate Expected Counts  {.smaller}

```{r}
#| echo: true
#| warning: false
#| message: false

library(tidyverse)
design_matrix <- expand.grid(
  intercept = 1,
  fem = 1,
  mar = 1,
  kid5 = 0,
  phd = mean(bioChemists$phd),
  ment = mean(bioChemists$ment)
) 

cat("The expected number of articles published, assuming the above  covariate values:\n",
   as.matrix(design_matrix) %*%  coef(my_model) |> exp() |> as.numeric(),
   "\narticles"
    )

```

## Generate Expected Counts  {.smaller}

```{r}
#| echo: true
#| warning: false
#| message: false
library(tidyverse)
design_matrix <- expand.grid(
  intercept = 1,
  fem = 0,
  mar = 1,
  kid5 = 1,
  phd = mean(bioChemists$phd),
  ment = mean(bioChemists$ment)
) 

cat("The expected number of articles published, assuming the abovecovariate values:\n",
   as.matrix(design_matrix) %*%  coef(my_model) |> exp() |> as.numeric(),
   "\narticles"
    )

```

## Simulate Uncertainty {.smaller}

```{r}
#| echo: true
#| warning: false
#| message: false
#| 
library(dplyr)

library(tidyverse)
design_matrix <- expand.grid(
  intercept = 1,
  fem = 0,
  mar = 1,
  kid5 = 2,
  phd = mean(bioChemists$phd),
  ment = mean(bioChemists$ment)
) 

parameter_draws = MASS::mvrnorm(1000, mu = coef(my_model), Sigma = vcov(my_model))
sampled_values = as.matrix(design_matrix) %*%  t(parameter_draws) |> exp() |> t() |> as.tibble()

sampled_values |>
  summarize(
    mean = mean(V1),
    lower = quantile(V1, 0.025),
    upper = quantile(V1, 0.975)
  )
```


## Simulate Uncertainty {.smaller}

- And varying multiple variables simultaneously, predicting what happens in different combinations of covariate values.

- Note that the code is identical, the exception being `bind_cols`. This is just to label the prediction groups in the final output.

```{r}
#| echo: true
#| warning: false
#| message: false

library(tidyverse)
design_matrix <- expand.grid(
  intercept = 1,
  fem = c(0,1),
  mar = c(0,1),
  kid5 = c(0,1,2),
  phd = mean(bioChemists$phd),
  ment = mean(bioChemists$ment)
) 

parameter_draws = MASS::mvrnorm(1000, mu = coef(my_model), Sigma = vcov(my_model))
sampled_values = as.matrix(design_matrix) %*%  t(parameter_draws) |> exp() |> as_tibble()

sampled_values_with_groups <- bind_cols(
  design_matrix |> select(fem, mar, kid5), 
  sampled_values
)

sampled_values_with_groups |>
  pivot_longer(cols = starts_with("V"), names_to = "draw", values_to = "predicted") |>
  group_by(fem, mar, kid5) |>
  summarize(
    mean = mean(predicted),
    lower = quantile(predicted, 0.025),
    upper = quantile(predicted, 0.975),
    .groups = "drop"
  )
```
## Generating predictions {.smaller}

- Use the Poisson CDF in `R` to generate predicted probabilities of counts.

- We can use the same code, just change `exp()` to `ppois()`

## The Negative Binomial Regression Model  {.smaller}

-   The NBRM stems from the negative binomial distribution

-   Recall the binomial density is the PDF stemming from $k$ independent bernoulli trials. The $\theta$ parameter will govern its shape.

-    We can modify the logic (and code) slightly to generate a probability of observing $r$ successes, given $n$ trials.

```{r}
#| echo: true
#| warning: false
#| message: false

library(tidyverse)
design_matrix <- expand.grid(
  intercept = 1,
  fem = c(0,1),
  mar = c(0,1),
  kid5 = c(0,1,2),
  phd = mean(bioChemists$phd),
  ment = mean(bioChemists$ment)
) 

parameter_draws = MASS::mvrnorm(1000, mu = coef(my_model), Sigma = vcov(my_model))
sampled_values = as.matrix(design_matrix) %*%  t(parameter_draws) |> exp() |> ppois(2) |> as_tibble()

sampled_values_groups <- bind_cols(
  design_matrix |> select(fem, mar, kid5), 
  sampled_values
)

sampled_values_groups |>
  pivot_longer(cols = starts_with("V"), names_to = "draw", values_to = "predicted") |>
  group_by(fem, mar, kid5) |>
  summarize(
    mean = mean(predicted),
    lower = quantile(predicted, 0.025),
    upper = quantile(predicted, 0.975),
    .groups = "drop"
  )
```
## The Negative Binomial Regresion {.smaller}

- In **negative binomial regression**, count data are modeled with overdispersion:

- **Mean**: $\mu = \exp(X\beta)$ 
- **Variance**: $\mu + \frac{\mu^2}{\theta}$ (larger than Poisson's $\mu$)

- `glm(formula, data = data, family = quasipoisson(link = "log"))`

## The Negative Binomial Regresion {.smaller}

```{r}
#| echo: true

glm(art~kid5 + fem + mar + phd + ment , 
            data=bioChemists,
            family=quasipoisson(link="log")) |>
  summary()

```

## Capturing Dispersion  {.smaller}

-  If we have overdispersion ($E(y)<var(y)$) then standard errors will be too small and we will be too over confident in our results.
- Instead, let's assume that the rate parameter is subject to error.
- That is, it follows some distribution. We'll use Gamma.
- The Negative Binomial Distribution is a mixture between a Poisson and a Gamma distribution.


## Capturing Dispersion: The Gamma-Poisson Mixture  {.smaller}

**The Problem**: If we have overdispersion ($E(y) < \text{Var}(y)$), standard errors will be too small and we will express too much confidence in our results.

- The **Negative Binomial**: $\text{Var}(Y_i) = \mu_i + \frac{\mu_i^2}{\theta}$ (variance > mean)

- $\theta$ = **dispersion parameter** 
- As $\theta \to \infty$: variance approaches Poisson.
- As $\theta \to 0$: high overdispersion.

$$
\lambda_i \sim \text{Gamma}(\text{shape}=\theta, \text{rate}=\frac{\theta}{\mu_i})
$$
-  Given this updated rate, then $Y_i$ just follows the Poisson.

$$Y_i|\lambda_i \sim \text{Poisson}(\lambda_i \cdot e_i)$$
$$\mu_i=exp(\alpha+\sum_K \beta_k x_{k,i})$$

## Application: Negative Binomial Regression {.smaller}

- The coefficients are the same as the Poisson regression.

- The general approach to extract meaningful estimates is also the same.

- The only difference is that we use `family=quasipoisson(link="log")` in `glm()`

- And estimate a $\theta$ parameter that captures overdispersion.

- The standard errors change.

## {.smaller}

**Poisson Regression Model**

```{r}
#| echo: true
#| warning: false
#| message: false

library(pscl)
data(bioChemists)
bioChemists$fem<-as.numeric(bioChemists$fem)-1
bioChemists$mar<-as.numeric(bioChemists$mar)-1
my_model = glm(art~kid5 + fem + mar + phd + ment , 
            data=bioChemists,
            family=poisson(link="log"))
summary(my_model)
```

## {.smaller}

**Negative Binomial Regression Model**

```{r}
#| echo: true
#| warning: false
#| message: false

library(pscl)
data(bioChemists)
bioChemists$fem<-as.numeric(bioChemists$fem)-1
bioChemists$mar<-as.numeric(bioChemists$mar)-1
my_model = glm(art~kid5 + fem + mar + phd + ment , 
            data=bioChemists,
            family=quasipoisson(link="log"))
summary(my_model)
```

